Setting parameters for biological models with ANIMO

### 2014-02-03 Paper.tex
Added some future work with references to other t.a. and hybrid automta works

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{SynCoP 2014} % Name of the event you are submitting to
\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage[strings]{underscore}

@@

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=.59\textwidth]{images/animo-ui}
  \caption{ANIMO user interface.\label{fig:animo-ui}}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=.4\textwidth]{images/animo-edge-ui}
  \caption{Editing the parameters for an ANIMO interaction.\label{fig:animo-edge-ui}}
\end{center}
\end{figure}

@@

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
Thanks to parameter sweep, we can make a more exhaustive search than the one presented in Figure~\ref{fig:small-example-feedback},
using an automatic approach instead of a manual trial-and-error process.
A logarithmic parameter sweep was performed, letting the parameter values
vary over the 5 pre-sets ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), ``medium'' ($k = 0.004$), ``fast'' ($k = 0.008$), ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $\mbox{\it parameter choices}^{\mbox{\it\scriptsize interactions}} = 5 ^ 5 = 3\,125$ simulations, that were computed
in 43 seconds %43 s with 16 threads. It is 57 with 8 threads (cpu is exploited less than 100%)
on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The model we consider here is small enough for one simulation to take
less than a second, but the exponential growth of the number of simulations required for a parameter sweep
can make the sheer number of tasks challenging for any processing unit.
% Indeed, a more thorough
% parameter sweep that considers 10 different setting for each of the 5 interactions of the considered
% model would require the computation of $10^5 = 100\,000$ simulations.
For example, a more complex model involving 10 interactions instead of 5 would require $5^{10} = 9\,765\,625$ simulations,
which could be computed in approximately a day and a half on an eight-core machine.
Still, the possibility to compute simulations in parallel provides a significant help in reducing
the impact of an exponential growth of the space to be explored.
% Yet, such an
% estimation would still be better than the twelve days it would take if the simulations were to be computed in sequence.
% the same experiment
% was run executing only one simulation at a time, and it took about 7 minutes and 34 seconds, %453837 ms
% which is in line with the fact that this type of analysis is trivially parallelizable.

The result window shown by ANIMO at the end of the computation of the $3\,125$ simulations can
be seen in Figure~\ref{fig:parameter-sweep-result}, where the best 6 results are listed in form of graphs.
By interacting with the interface, the user can choose to see more of the top-scoring results: this will allow to check
the width of the parameter settings for which the wanted behavior can still be considered ``close enough'' to
the experimental data. Pressing one of the buttons labelled ``I want this'' under one of the graphs
allows the user to copy the parameters used in the selected simulation to the network model.

@@

% dopo aver presentato gli strumenti che forniamo, definiamo una procedura consigliata per usare ANIMO nel modo più efficace.
% Ovviamente, oltre alle best practices serve anche l'esperienza, sia intesa come conoscenza di biologia che come esperienza
% nel campo della modellazione. Nota che non serve comunque conoscenza di T.A., visto che rimangono sempre nascosti dietro la GUI
% Esempio/case study!
\section{Suggested ANIMO workflow}
In order to better profit from what ANIMO offers to the biologist, we suggest a step-by-step modeling workflow:
\begin{enumerate}
  \item\label{step:def-topology} Define a starting topology, based on literature data and current knowledge on the network.
  \item\label{step:set-rates} Choose the interaction parameters from the pre-set qualitative values,
    based on experience (e.g., protein expression is much slower than phosphorylation).
    The choice on the approximated kinetics can be for the most part based on scenario 1, referring to Section~\ref{sec:parameter-editing}.
  \item\label{step:check-behavior} Check that the network behaves as expected (e.g. addition of EGF makes ERK activity rapidly increase and successively decrease).
  \item Possibly iterate the steps~\ref{step:def-topology} to~\ref{step:check-behavior} until the network behaves as expected.
  \item Compare the model with experimental data.
  \item\label{step:precise-fit} If the model does not fit the data, change the parameters using more precise numerical values and/or scenario settings.
  \item\label{step:param-sweep} If the data fitting is still not satisfactory, apply parameter sweep to (subsets of) the model as described in Section~\ref{sec:parameter-sweep}.
  \item When a good fit is found, add non-determinism to the model and ensure that the
     behavior does not change significantly under reasonably high uncertainty settings.
  \item If the model can fit the data only with very precise parameter settings, or a
    parameter configuration that fits the data cannot be found after a reasonably extensive search
    on the parameter space, change the network topology possibly adding more nodes from literature.
    If no candidates can be found in literature, speculative candidates can be found by ``playing'' with the current
    configuration in ANIMO's user interface.
    We advise to let expert biologists perform this step on their own, or to perform it under their supervision,
    in order to keep the network both simple and realistic. For the same reason, we also advise not to introduce more than a couple of nodes at a time.
  \item Parameters for newly introduced interactions can be chosen as in step~\ref{step:set-rates}, possibly iterating
    steps~\ref{step:def-topology} to~\ref{step:param-sweep} to refine the parameters.
  \item To check how changes in the network are reflected in the behavior of the model, we advise to use the comparison tools
    described in Section~\ref{sec:comparing-versions}.
  \item When a model is obtained that behaves more closely to the experimental data,
  its fitness can be improved going back to step~\ref{step:precise-fit}.
\end{enumerate}
If the model fits the data reasonably well only after modifications to the initial topology, it may contain some hypothetical or unexpected nodes.
In that case, future experiments can be derived directly from the hypotheses introduced in the model and
help the biologists decide which are the most precise explanations for the experimental observations.

In general, the process described above allows the biologists to formalize their knowledge,
and makes it easier to have group discussions when trying to combine multiple subnetworks in a more comprehensive model.
It is worth noting that experience in both the specific biological setting and generic modeling of biological
events represents an important asset for a faster and more realistic modeling process with ANIMO. However, as
we claimed with all previous versions of the tool, it is still the case that no experience in formal methods
is needed in order to use ANIMO. In particular, an ideal ANIMO user could have no experience on the \tas\ foundations
of the tool and yet fully profit from its features.

% anche Related work: es. Copasi
% Altro su parameter sweep/sensitivity analysis? Noi usiamo un approccio un po' naif, senza tante pretese, senza tanti numeri.
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri
\section{Conclusions}
ANIMO has been extended with a proper support for parameter synthesis. Now the tool can be used more efficiently
to model biological networks, without concentrating exclusively on their topology.
In particular, the addition of an automatic parameter sweep feature allows the user
to save considerable amounts of time, keeping closer to experimental data without the disadvantages
coming from a hand-made parameter search.

>>>>

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{SynCoP 2014} % Name of the event you are submitting to
\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage[strings]{underscore}
\usepackage[font=scriptsize,labelfont={bf}]{caption}
\usepackage[justification=raggedright,nearskip=10pt,farskip=0pt]{subfig}

@@

\begin{figure}[htbp]
\begin{center}
\subfloat[\label{fig:animo-ui}]{\includegraphics[width=.55\textwidth]{images/animo-ui}} \quad
\subfloat[\label{fig:animo-edge-ui}]{\includegraphics[width=.4\textwidth]{images/animo-edge-ui}}
\caption{ANIMO user interface. (a) The main window of Cytoscape, with the ANIMO
plug-in showing a model network and the result of a simulation. (b) Editing the parameters for an ANIMO interaction.}
\end{center}
\end{figure}

% \begin{figure}[htbp]
% \begin{center}
%   \includegraphics[width=.4\textwidth]{images/animo-edge-ui}
%   \caption{Editing the parameters for an ANIMO interaction.\label{fig:animo-edge-ui}}
% \end{center}
% \end{figure}

@@

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
Thanks to parameter sweep, we can make a more exhaustive search than the one presented in Figure~\ref{fig:small-example-feedback},
using an automatic approach instead of a manual trial-and-error process.
We started by choosing scenario 2 for the two interactions involving ERK, while the other three interactions were left to scenario 1:
this allows us to obtain a smoother looking graph for ERK.
A logarithmic parameter sweep was performed, letting the parameter values
vary over the 5 pre-sets ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), ``medium'' ($k = 0.004$), ``fast'' ($k = 0.008$), ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $\mbox{\it parameter choices}^{\mbox{\it\scriptsize interactions}} = 5 ^ 5 = 3\,125$ simulations, that were computed
in 43 seconds %43 s with 16 threads. It is 57 with 8 threads (cpu is exploited less than 100%)
on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The model we consider here is small enough for one simulation to take
less than a second, but the exponential growth of the number of simulations required for a parameter sweep
can make the sheer number of tasks challenging for any processing unit.
% Indeed, a more thorough
% parameter sweep that considers 10 different setting for each of the 5 interactions of the considered
% model would require the computation of $10^5 = 100\,000$ simulations.
For example, a more complex model involving 10 interactions instead of 5 would require $5^{10} = 9\,765\,625$ simulations,
which could be computed in approximately a day and a half on an eight-core machine.
Still, the possibility to compute simulations in parallel provides a significant help in reducing
the impact of an exponential growth of the space to be explored.
% Yet, such an
% estimation would still be better than the twelve days it would take if the simulations were to be computed in sequence.
% the same experiment
% was run executing only one simulation at a time, and it took about 7 minutes and 34 seconds, %453837 ms
% which is in line with the fact that this type of analysis is trivially parallelizable.

The result window shown by ANIMO at the end of the computation of the $3\,125$ simulations can
be seen in Figure~\ref{fig:parameter-sweep-result}, where the best 6 results are listed in form of graphs.
By interacting with the interface, the user can choose to see more of the top-scoring results: this will allow to check
the width of the parameter settings for which the wanted behavior can still be considered ``close enough'' to
the experimental data. Pressing one of the buttons labelled ``I want this'' under one of the graphs
allows the user to copy the parameters used in the selected simulation to the network model.
The graph in Figure~\ref{fig:animo-ui} was obtained by choosing the first graph shown in Figure~\ref{fig:parameter-sweep-result},
i.e. setting the parameters as in Table~\ref{tab:parameters}. It can be seen that the model does not
fit the data very precisely: this is because some nodes known to influence the activity of ERK
were purposely left out of the example for simplicity's sake.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|}
\hline
 Reaction & $k$ \\
\hline
\hline
EGF $\rightarrow$ EGFR & Fast \\
EGFR $\rightarrow$ EGFR internalization & Medium \\
EGFR internalization $\dashv$ EGFR & Very Fast \\
EGFR $\rightarrow$ ERK & Fast \\
ERK phosphatase $\dashv$ ERK & Medium\\
\hline
\end{tabular}
\caption{Qualitative values of the parameters $k$ used in the example model to obtain the graph shown in Figure~\ref{fig:animo-ui}.\label{tab:parameters}}
\end{table}

% dopo aver presentato gli strumenti che forniamo, definiamo una procedura consigliata per usare ANIMO nel modo più efficace.
% Ovviamente, oltre alle best practices serve anche l'esperienza, sia intesa come conoscenza di biologia che come esperienza
% nel campo della modellazione. Nota che non serve comunque conoscenza di T.A., visto che rimangono sempre nascosti dietro la GUI
% Esempio/case study!
\section{Suggested ANIMO workflow}
In order to better profit from what ANIMO offers to the biologist, we suggest a step-by-step modeling workflow:
\begin{enumerate}
  \item\label{step:def-topology} Define a starting topology, based on literature data and current knowledge on the network.
  \item\label{step:set-rates} Choose the interaction parameters from the pre-set qualitative values,
    based on experience (e.g., protein expression is much slower than phosphorylation).
    The choice on the approximated kinetics can be for the most part based on scenario 1, referring to Section~\ref{sec:parameter-editing}.
  \item\label{step:check-behavior} Check that the network behaves as expected (e.g. addition of EGF makes ERK activity rapidly increase and successively decrease).
  \item Possibly iterate the steps~\ref{step:def-topology} to~\ref{step:check-behavior} until the network behaves as expected.
  \item Compare the model with experimental data.
  \item\label{step:precise-fit} If the model does not fit the data, change the parameters using more precise numerical values and/or scenario settings.
  \item\label{step:param-sweep} If the data fitting is still not satisfactory, apply parameter sweep to (subsets of) the model as described in Section~\ref{sec:parameter-sweep}.
  \item When a good fit is found, add non-determinism to the model and ensure that the
     behavior does not change significantly under reasonably high uncertainty settings.
  \item If the model can fit the data only with very precise parameter settings, or a
    parameter configuration that fits the data cannot be found after a reasonably extensive search
    on the parameter space, change the network topology possibly adding more nodes from literature.
    If no candidates can be found in literature, speculative candidates can be found by ``playing'' with the current
    configuration in ANIMO's user interface.
    We advise to let expert biologists perform this step on their own, or to perform it under their supervision,
    in order to keep the network both simple and realistic. For the same reason, we also advise not to introduce more than a couple of nodes at a time.
  \item Parameters for newly introduced interactions can be chosen as in step~\ref{step:set-rates}, possibly iterating
    steps~\ref{step:set-rates} to~\ref{step:param-sweep} to refine the parameters.
  \item To check how changes in the network are reflected in the behavior of the model, we advise to use the comparison tools
    described in Section~\ref{sec:comparing-versions}.
\end{enumerate}
If the model fits the data reasonably well only after modifications to the initial topology, it may contain some hypothetical or unexpected nodes.
In that case, future experiments can be derived directly from the hypotheses introduced in the model, and
help the biologists decide which are the most precise explanations for the experimental observations.

A simplified graphical representation of the workflow can be found in Figure~\ref{fig:workflow}: the three
degrees of precision with the parameter settings are highlighted with progressively darker shades of blue.
Moreover, the figure helps to stress the fact that in our method changes in topology (green boxes) should be made as a
last step, after a reasonable attempt at making the model fit via parameter adjustments.
Again, it is important to remember that if a model behaves correctly only for very narrow parameter settings
then it is not as robust as expected, so its topology may need to be reviewed.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{images/workflow}
\caption{ANIMO modeling workflow.\label{fig:workflow}}
\end{figure}

The process described above allows the biologists to formalize their knowledge,
and makes it easier to have group discussions when trying to combine multiple subnetworks in a more comprehensive model.
It is worth noting that experience in both the specific biological setting and generic modeling of biological
events represents an important asset for a faster and more realistic modeling process with ANIMO. However, as
we claimed with all previous versions of the tool, it is still the case that no experience in formal methods
is needed in order to use ANIMO. In particular, an ideal ANIMO user could have no experience on the \tas\ foundations
of the tool and yet fully profit from its features.

% anche Related work: es. Copasi
% Altro su parameter sweep/sensitivity analysis? Noi usiamo un approccio un po' naif, senza tante pretese, senza tanti numeri.
%parametric model checking, hytech?
%imitator II ??
\section{Conclusions}
ANIMO has been extended with a proper support for parameter synthesis. Now the tool can be used more efficiently
to model biological networks, without concentrating exclusively on their topology.
In particular, the addition of an automatic parameter sweep feature allows the user
to save considerable amounts of time, keeping closer to experimental data without the need
to perform a hand-made parameter search.

In the future, we plan to improve our parameter analysis techniques, possibly using some of
the techniques already developed for testing parameter sensitivity~\cite{inverse-ta1, inverse-ta2} and
robustness~\cite{robust-ta1, robust-ta2, robust-ta3, robust-ta4} in \tas\ and hybrid systems more in general~\cite{decidability-ta-ha, imitator2}.
As already stated, parameter sensitivity is very important in biological models: if a treatment
derived from a model is based on narrow bounds for a parameter, that parameter must not exhibit significant
variety among patients. By introducing an automatic analysis for parameter sensitivity, our tool
could point out the zones of a network model where the user should concentrate more, for example by
improving the topology, or by more focused parameter sweeps.
Moreover, we also aim at applying some results from automata learning to the generation of
\tas\ models based on experimental data: generating models that exhibit specific behaviors
implies a careful definition of their parameter space.
Finally, parametric model checking represents an interesting development direction:
having a tool automatically output the parameter intervals for which some
given outputs are observed is a highly valued resource for the biological researcher.
Investigating the properties of the specific \tas\ models used in ANIMO
could lead to important tractability results valid in our case, allowing us
to apply methods not valid for \tas\ in general.
%FUTURE WORK!!
%parameter sensitivity: se un trattamento dipende da un parametro specifico, e' il caso che non si registri troppa varieta' per quel parametro tra i pazienti.
% per quei parametri non sensibili, si eviti di concentrare troppi sforzi. Il tool potrebbe automaticamente determinare le ``zone calde'' della rete
%anche alla luce di una possibile generazione di modelli t.a. basati su dati sperimentali, trovare modi intelligenti di generare i parametri
%il fatto che i modelli che usiamo possano identificare un sottinsieme ristretto di t.a. potrebbe essere vantaggioso per la trattabilita' dei problemi
%parametrized model checking: con che parametri e' possibile raggiungere questo stato?
%definisci un obiettivo di quello che vuoi ottenere, e lui ti dice i [range di] parametri che arrivano a quell'evoluzione
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri
